---
title: "Data Mining Project - Arcene"
author: "Suman Gorkhali"
date: "2026-03-30"
output: html_document
---


```{r library}
library(foreign)
library(randomForest)
library(RRF)
library(tree)
library(glmnet)
library(MASS)

set.seed(20551088)
```

```{r arceneData}
arcene = read.arff("arcene_data.arff")
arcene$Class = ifelse(arcene$Class == 2, 0, 1)
arcene$Class = as.factor(arcene$Class)

train_idx = as.integer(x = sample(nrow(arcene), size = 0.8 * nrow(arcene)))

# dataset to store model stat
model_stat = data.frame(
                        model = character(),
                        Sensitivity = numeric(),
                        Specificity = numeric(),
                        Error = numeric(),
                        stringsAsFactors = FALSE
                        )

```

```{r testAccuracyFunction}
testAccuracy <- function(prediction, k){
  truth_table = table(Prediction = prediction, Actual = arcene[-train_idx, "Class"])
  print(truth_table)
  
  TP = truth_table["1","1"]
  TN = truth_table["0","0"]
  FP = truth_table["1","0"]
  FN = truth_table["0","1"]
  
  sensitivity = TP/ (TP + FN)
  
  specificity = TN/(TN + FP)

  error = (FP + FN)/(FP + FN + TP + TN)
  
  # store in the model_stat data frame
  rowNum = (dim(model_stat)[1] + 1)
  
  model_stat[rowNum, "model"] <<- k
  model_stat[rowNum, "Sensitivity"] <<- sensitivity
  model_stat[rowNum, "Specificity"] <<- specificity
  model_stat[rowNum, "Error"] <<- error
  
  return(c(sensitivity = paste(k,' :', round(sensitivity,2)), 
           specificity = paste(k, ': ', round(specificity,2)), 
           error = paste(k, ': ', round(error,2))
           )
         )
}

```


# ```{r lasso}
# set.seed(20551088)
# 
# x = model.matrix(Class ~ ., data = arcene)[, -1]
# y = arcene$Class 
# 
# fitLasso = cv.glmnet(x[train_idx, ], y[train_idx], alpha = 1, family = "binomial")
# 
# reducedDim = coef(fitLasso, s = 'lambda.min')
# reducedDim = data.frame(as.matrix(reducedDim))
# index = which(reducedDim$lambda.min != 0)
# lasso_Dimension = rownames(reducedDim)[index]
# lasso_Dimension
# length(lasso_Dimension)
# 
# predLasso_prob = predict(fitLasso, newx = x[-train_idx, ], s = "lambda.min", type = "response")
# predLasso = ifelse(predLasso_prob > 0.5, 1, 0)
# 
# plot(fitLasso)
# 
# lassoTestAccuracy = testAccuracy(predLasso, 'lasso') 
# lassoTestAccuracy
# 
# ```

```{r pca}
#PCA + Logistic Regression
set.seed(20551088)

#  predictor matrix and response
x <- model.matrix(Class ~ ., data = arcene)[, -1]
y <- arcene$Class


x.train <- x[train_idx, ]
y.train  <-y[train_idx]

x.valid <- x[-train_idx, ]
y.valid <- y[-train_idx]


nzv <- apply(x.train, 2, sd) != 0
x.train <- x.train[, nzv]
x.valid <- x.valid[, nzv]



# PCA
pca <- prcomp(x.train, scale = TRUE)

# loadings
pca$rotation[1:50, 1:4]

# scores
head(pca$x)
pca$x[1:2, 1:2]

# variance explained
summary(pca)
sum(pca$sdev^2)
str(summary(pca))
summary(pca)$importance[3, ]   # cumulative proportion
summary(pca)$importance[2, ]   # proportion explained

# biplot
biplot(pca, scale = 0)

# scree plot
plot(pca)

# proportion variance explained for first 10 PCs
plot(1:20, summary(pca)$importance[2, 1:20],
     type = "l",
     xlab = "Prin Comps",
     ylab = "Proportion Variance Explained",
     main = "Variance Explained by First 10 PCs")

# cumulative variance explained for first 10 PCs
plot(1:20, summary(pca)$importance[3, 1:20],
     type = "l",
     xlab = "Prin Comps",
     ylab = "Cumulative Proportion",
     main = "Cumulative Variance Explained")

# PCA scores plot
plot(pca$x[,1], pca$x[,2],
     col = as.numeric(y.train),
     pch = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA Scores Plot")

legend("topright",
       legend = levels(y.train),
       col = 1:length(levels(y.train)),
       pch = 19)

# choose number of PCs K= 20


# training PC scores
train.pcs <- as.data.frame(pca$x[, 1:20])
train.pcs$Class <- y.train

# validation PC scores
valid.scores <- predict(pca, newdata = x.valid)
valid.pcs <- as.data.frame(valid.scores[, 1:20])
valid.pcs$Class <- y.valid
```

```{r pcaTHENglm}
set.seed(20551088)
fit.log <- glm(Class ~ ., data = train.pcs, family = binomial)

summary(fit.log)

# prediction
pred.log.prob <- predict(fit.log, newdata = valid.pcs, type = "response")
pred.log  <- ifelse(pred.log.prob > 0.5, 1, 0)
pred.log  <-as.factor(pred.log)

# results
table(pred.log, y.valid)

pcaLogTestAccuracy <- testAccuracy(pred.log, "pca + fit.log")
pcaLogTestAccuracy
```



# ```{r cvtree}
# set.seed(20551088)
# fitTree = tree(as.factor(Class) ~., data = arcene, subset = train_idx)
# #plot(fitTree, cex = 0.01)
# 
# cvTree = cv.tree(fitTree)
# plot(cvTree$size, cvTree$dev)
# 
# prunedTree = prune.tree(fitTree, best = 3, method = 'deviance')
# plot(prunedTree)
# text(prunedTree, cex = 0.5)
# 
# predPrunedTree = predict(prunedTree, arcene[-train_idx,], type = 'class')
# summary(prunedTree)
# 
# treeTestAccuracy = testAccuracy(predPrunedTree, 'prunedTree')
# treeTestAccuracy
# 
# 
# ```


```{r Bagging}
set.seed(20551088)
fitBag = randomForest(Class ~., data = arcene, subset = train_idx, mtry = 10000)
predBag = predict(fitBag, arcene[-train_idx,])

truth_table = table(Prediction = predBag, Actual = arcene[-train_idx, "Class"])

plot(fitBag)

baggingTestAccuracy = testAccuracy(predBag, 'Bagging')
baggingTestAccuracy


```


```{r VariableImportance}
set.seed(20551088)
fitVI = randomForest(Class ~., data = arcene, subset = train_idx,
                     importance = TRUE, ntree = 400)
plot(fitVI, main = 'Random Forest with Variable Importance')
#randomForest::importance(fitVI)
randomForest::varImpPlot(fitVI, main = 'Random Forest with Variable Importance')

predVI = predict(fitVI, arcene[-train_idx,])

baggingTestAccuracy = testAccuracy(predVI, 'randomForest Variable Importance')
baggingTestAccuracy

```




```{r GRRF}
set.seed(1)
# Guided Regularized Random Forest
fit_rrf = RRF(arcene[train_idx, -10001], arcene[train_idx, 10001], flagReg = 0, ntree = 500)
importance = fit_rrf$importance[,1]

# Hyperparameter
gamma = 0.75

# Guided RRF coefficients
coefGRRF = (1 - gamma) + gamma * (importance/(max(importance)))

# Guided Regularized Random Forest
fit_guided_rrf = RRF(arcene[train_idx, -10001], arcene[train_idx, 10001], flagReg = 1, coefReg = coefGRRF)

# Selected Features
selectedFeatures = fit_guided_rrf$feaSet
length(selectedFeatures)

# Prediction
predGRRF = predict(fit_guided_rrf, arcene[-train_idx,])

GRRFtestAccuracy = testAccuracy(predGRRF, 'GRRF')
GRRFtestAccuracy



```

```{r GRRFmultipleGamma}
# Try multiple gamma values
# gamma_values = c(0, 0.25, 0.5, 0.75, 1)
# 
# results = data.frame()
# 
# for(g in gamma_values){
#   coefGRRF = (1 - g) + g * (importance/(max(importance)))
#   
#   fit_temp = RRF(arcene[train_idx, -10001],
#                  arcene[train_idx, 10001],
#                  flagReg = 1,
#                  coefReg = coefGRRF)
#   
#   pred = predict(fit_temp, arcene[-train_idx,])
#   
#   acc = testAccuracy(pred, paste("GRRF_gamma_", g))
# }
```

## LDA

LDA assumes both classes share the same covariance matrix and finds
the linear combination that best separates them. After PCA reduction to
130 components, the covariance matrix is invertible and we can fit the
model.



```{r LDA}
set.seed(20551088)

train.x <- arcene[train_idx, -ncol(arcene)]
test.x <- arcene[-train_idx, -ncol(arcene)]
train.y <- arcene$Class[train_idx]
test.y <- arcene$Class[-train_idx]

# remove zero-variance columns
keep <- apply(train.x, 2, sd) > 0
train.x <- train.x[, keep]
test.x <- test.x[, keep]
cat("Removed", sum(!keep), "zero-variance columns,", sum(keep), "remaining\n")

pca.fit <- prcomp(train.x, scale. = TRUE)

# how many components for different variance thresholds
cum.var <- cumsum(pca.fit$sdev^2) / sum(pca.fit$sdev^2)
cat("Components for 80% variance:", which(cum.var >= 0.80)[1], "\n")

cat("Components for 90% variance:", which(cum.var >= 0.90)[1], "\n")

cat("Components for 95% variance:", which(cum.var >= 0.95)[1], "\n")

cat("Components for 99.99% variance:", which(cum.var >= 0.9999)[1], "\n")


ncomp.95 <- which(cum.var >= 0.95)[1]

train.pca <- data.frame(pca.fit$x[, 1:ncomp.95], Class = train.y)
test.pca <- data.frame(predict(pca.fit, test.x)[, 1:ncomp.95])

fit.lda <- lda(Class ~ ., data = train.pca)

pred.lda <- predict(fit.lda, test.pca)
lda.metrics <- testAccuracy(pred.lda$class, "pca + LDA")

lda.metrics
```


### Discriminant Score Distribution

Looking at how well LD1 separates the two classes on the training
data.

```{r chunk16}
lda.train.scores <- predict(fit.lda, train.pca)

scores.0 <- lda.train.scores$x[train.y == "0"]
scores.1 <- lda.train.scores$x[train.y == "1"]

hist(scores.0, col = rgb(1, 0, 0, 0.4), xlim = range(lda.train.scores$x),
     breaks = 15, main = "LDA Discriminant Scores by Class", xlab = "LD1")
hist(scores.1, col = rgb(0, 0, 1, 0.4), breaks = 15, add = TRUE)
legend("topright", legend = c("Normal (0)", "Cancer (1)"),
       fill = c(rgb(1, 0, 0, 0.4), rgb(0, 0, 1, 0.4)))
```

### Posterior Probability Analysis

LDA gives each test observation a posterior probability of being
cancer. Points above the 0.5 line get classified as cancer.

```{r chunk17}
lda.test.post <- predict(fit.lda, test.pca)$posterior[, "1"]

plot(lda.test.post, col = as.numeric(test.y) + 1, pch = 19,
     ylab = "P(Cancer | X)", xlab = "Test Observation",
     main = "LDA Posterior Probabilities")
abline(h = 0.5, lty = 2)
legend("topleft", legend = c("Normal (0)", "Cancer (1)"), col = c(2, 3), pch = 19)
```

### Threshold Analysis

Lowering the threshold below 0.5 catches more cancer cases but also
increases false positives.

```{r chunk18}
thresholds <- seq(0.1, 0.9, by = 0.05)
thresh.sens <- rep(0, length(thresholds))
thresh.spec <- rep(0, length(thresholds))

for (j in 1:length(thresholds)) {
  pred.t <- ifelse(lda.test.post > thresholds[j], "1", "0")
  pred.t <- factor(pred.t, levels = c("0", "1"))
  cm <- table(pred.t, test.y)
  if (nrow(cm) == 2 && ncol(cm) == 2) {
    thresh.sens[j] <- cm["1", "1"] / sum(cm[, "1"])
    thresh.spec[j] <- cm["0", "0"] / sum(cm[, "0"])
  }
}

plot(thresholds, thresh.sens, type = "l", col = "blue", lwd = 2,
     ylim = c(0, 1), xlab = "Threshold", ylab = "Rate",
     main = "LDA: Sensitivity vs Specificity by Threshold")
lines(thresholds, thresh.spec, col = "red", lwd = 2)
legend("right", legend = c("Sensitivity", "Specificity"),
       col = c("blue", "red"), lwd = 2)
```


```{r Comparative}
library(ggplot2)
library(tidyr)
library(dplyr)


plot_model_stat <- function(data) {
  df_long <- data %>%
    pivot_longer(cols = -model, names_to = "metric", values_to = "value")

  ggplot(df_long, aes(x = metric, y = value, fill = model)) +
    geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.2) +

    scale_fill_brewer(palette = "Set2") +
    theme_minimal() +

    labs(
      title = "Comparative Model Performance Stat",
     y = "Probability Score",
      fill = "Model:  "
    ) +
     # Add a clean theme
    theme(
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank()
    )
}

plot_model_stat(model_stat)
```



